---
title: "Object-Oriented Programming in R"
author: Christopher Eeles
date: 2021-01-27
output: ioslides_presentation
---

# What is Object-Oriented Programming?

## Object-Oriented Programming

- Programming paradigm that organizes data and methods into classes
- One way to organize your code to make it easier to understand, reason
about, and most importantly: reuse!

## The Four Principles of OOP

- Encapsulation: objects manage their own state via methods
- Abstraction: hide complexity of implementation; only expose interfaces
- Inheritance: allows similar objects to share properties and methods
- Polymorphism: same method names, different method implementations

## Why Do I Care?

- OOP can help group related data into a standard format
- OOP allows you to make valid (and guaranteed) assumptions about the structure
of your data
- OOP allows you to build on the work of others by reusing or extending existing
classes
- OOP allows you to write less code!
- OOP can abstract away input validation and reduce boiler plate error checking
such as input validation
- Interfaces are awesome (more on this later)

# Generic Function OOP

## What is Generic Function OOP?

- In generic function OOP, methods are first class objects
- They are NOT included in classes
- This allows modelling classes to be independent of modelling behaviour
    - What a thing is does not limit what it does
    - Much like in English, nouns are independent of verbs!

## Generic Function OOP vs Message-Passing OOP

- The OOP system in R is very different from traditional message-passing OOP in
languages like C++ or Java
- The methods available to a class are not in the definition of that class and
are not set at compile time (indeed, in R there is no compile time)
- Instead, method dispatch is used to decide which implementation to call at
runtime
- If there is no method implemented for that class; a method dispatch error
is returned letting the user know

## Benefits of Generic Function OOP

- In message-passing OOP, it is very easy to reuse nouns (classes)
- Consider a case where you like a class, but dislike the method implementation
- How is this resolved? Need to inherit from the class you like, define new
methods using polymorphism; now you have a whole new class you have to maintain...
- A better solution: just define new verbs for the noun! This is possible
in Generic Function OOP


# Bioconductor

## Bioconductor
- Bioconductor provides tools for the analysis and comprehension of high-throughput genomic data
  - This is where most of our R packages live! It is a great community with a ton
  of existing functionality to reuse
- The Bioconductor organization provides a rich set of S4 classes as core
infrastructure in the Bioconductor development community!
  - If you weren't convinced on the value of S4 by any of my previous arguments,
  this one should be enough to change your mind

## Bioconductor Packages

- BiocGenerics: implements and exports S4 generics for a wide variety of base R
functions
    - If you are using the S4 system, you should check if a generic already
    exists here before defining your own
    - Establishes a standard set of verbs to operate on classes; equivalent
    operations of different classes should use the same verb
- S4Vectors: implementation of base R vectors in S4
    - Several abstract classes (interface definitions)


# Using the S4 Object System

## S4 Objects

- An S4 object is composed of a class name, one or more slots, and zero or
more parent classes
- Slots can be accessed using the `@` operator; but this should only be
used for development!
- S4 objects are just

## Defining a Class

- Inheritance is accomplished using the `contains` parameter
- The new class contains all the slots from the old class
- New slots can be added using the `slots` argument

## Defining a Class

```{r, eval=TRUE, include=TRUE, echo=FALSE, message=FALSE, warning=FALSE}
library(S4Vectors)
library(BiocGenerics)
library(SummarizedExperiment)
```

```{r}

# For documentation on this object
?setClass
#' @inherits SimpleList
#' @slot testSlot A `list` that does nothing but shows us how to defined
#'   a new slot!
#'
#' @md
#' @export
.SumExpList <- setClass('SumExpList', contains='SimpleList',
    slots=list(testSlot='list'))
```

## Defining a Constructor

- It is convention not to use the raw constructor object constructor, but instead
defined a new method with better documentation of how the object gets built
- This isn't usually an S4 method; it is just a plain old R function

## Defining a Constructor

```{r}
SumExpList <- function(...) {
    # make an object of the parent class
    simpleList <- SimpleList(...)

    # define the new requirements for your class
    mcols(simpleList)$class <-
        rep('SummarizedExperiment', length(as.list(...)))
    testSlot <- list('item'='This is a test slot!')

    # call your private constructor
    SElist <- .SumExpList(simpleList, testSlot=testSlot)
    return(SElist)
}
```


## Class Validity

- This is the method that allows you to guarantee assumptions about your data!
- Implements a limited kind of static typing; since you know the type of
some of your parameters, you don't need to check in your functions!


## Class Validity

```{r, message=FALSE}
?setValidity
setValidity('SumExpList', function(object) {
    isSummarizedExperiment <-
        unlist(lapply(object@listData, is, class='SummarizedExperiment'))
    if (!all(isSummarizedExperiment)) {
        message('Items at indexes: ', paste0(which(!isSummarizedExperiment), collapse=', '),
        ' are not SummarizedExperiments. All items in a SumExpList must be',
        'SummarizedExperiments!')
        return(FALSE)
    } else {
        return(TRUE)
    }
})
```

## Class Validity

```{r}

tryCatch({
    sumExpList <- SumExpList(list(item='Not a SummarizedExperiment!'))
}, error=function(e) message(e))

validSumExpList <- SumExpList(list(mySE=SummarizedExperiment()))

```

## Inheritted Slots

- Inherited slots fall back to the method of the parent class if no new
method has been set for the child class
- Lets you keep a lot of great Bioconductor functionality (specifically around
metadata) with no cost!

## Inherited Slots

```{r }

metadata(validSumExpList)

mcols(validSumExpList)

```

## But What Slots Does an Object Have?

```{r}
slotNames(validSumExpList)
```

## Generic Functions

- Generic functions are template method definitions; they establish a verb
that can be reused by others freely

- Generic functions are constrained only by the names of the arguments in the
generic definition
    - Once you name an argument in a generic, it cannot be renamed
    - Unless you include the `...` parameter in the generic definition no new
    parameters can be specified in implementations (bad practice)

## Defining a Generic

```{r}
?setGeneric
#'
#' @param object A `SumExpList` object to get the testSlot from.
#'
#' @export
setGeneric('testSlot', function(object, ...) standardGeneric('testSlot'))
```

## Method Dispatch

- Once a generic is defined, specific method implementations are defined
based on the types of the arguments
- It is possible to define a different implementation based on the types of
one or more parameters to the generic


## Defining Methods

```{r}
?setMethod
setMethod('testSlot', signature(object='SumExpList'), function(object) {
    return(object@testSlot)
})
```

## Class Unions

- Allow you to define methods that work on a set of non-related classes
- Can be a good way to share verbs between different classes without a huge
nested hierarchy of inheritance.

## Class Unions
```{r}
setClassUnion('SumExpList_or_SummarizedExperiment',
    members=c('SumExpList', 'SummarizedExperiment'))
```

## More to come!

- Method dispatch with
-


# References

## OOP

- freeCodeComp: How to explain object-oriented programming concepts to a 6-year-old (https://www.freecodecamp.org/news/object-oriented-programming-concepts-21bb035f7260/)
- educative: What is Object Oriented Programming? OOP Explained in Depth (https://www.educative.io/blog/object-oriented-programming)
- Wikipedia: Object-oriented programming (https://en.wikipedia.org/wiki/Object-oriented_programming)

## S4

- Advanced R: The S4 object system (http://adv-r.had.co.nz/S4.html)
- Hansen, K.D. R - S4 Classes and Methods (https://kasperdanielhansen.github.io/genbioconductor/html/R_S4.html)
- Pages, H. (2016). A quick overview of the S4 class system (https://www.bioconductor.org/packages/release/bioc/vignettes/S4Vectors/inst/doc/S4QuickOverview.pdf)

## Bioconductor

- BiocGenerics (https://bioconductor.org/packages/release/bioc/html/BiocGenerics.html)
- An Overview of the S4Vectors Package (https://bioconductor.org/packages/release/bioc/vignettes/S4Vectors/inst/doc/S4VectorsOverview.pdf)
-